---
title: "Reproducible Notebook"
output: html_document
date: "`r Sys.Date()`"
---

# Data Collection
## Load all required libraries for easy transitions
```{r}
library(osmdata)
library(sf)
library(tidyverse)
library(stplanr)
library(units)
library(ggplot2)
library(corrplot)
library(tmap)
library(ggcorrplot)
library(osrm)

```

## Find the city boundary
```{r}
# open the wiki features page https://wiki.openstreetmap.org/wiki/Map_features on your browser to ensure seamless data collection.

edi_bnd <- getbb("Edinburgh", format_out = "polygon")

```

My chosen variable are buildings, amenities, natural, highway intersections and office. This is supported by studies from Franke et al (2004), Frank et al., (2021) and Saelens et al., (2003)


## Finding the buildings data
```{r}

edi_buildings <- opq (edi_bnd) %>%
   add_osm_feature(key = "building", value = c("apartments","bungalow","dormitory","hotel","residential", "terrace", "commercial", "supermarket", "warehouse", "religious")) %>%
   osmdata_sf() #this last line is to make the output a simple feature object
 
 edi_buildings

```


## Load amenities data
```{r}
edi_amenities <- opq(edi_bnd) %>%
  add_osm_feature(key = "amenity", value = c("cafe","fast_food", "post_office", "community_centre","casino","cinema","atm", "kindergarten", "clinic")) %>%
  osmdata_sf()

edi_amenities

```


## Load Natural environment data
```{r}
edi_natural <- opq(edi_bnd) %>%
  add_osm_feature(key = "natural", value = c("scrub","tree","shrubbery","tree_row")) %>%
  osmdata_sf()

edi_natural
```

## Load Intersections: highway data

```{r}
edi_highway <- opq(edi_bnd) %>%
  add_osm_feature(key = "highway", value = c("living_street", "crossing", "service","pedestrian","footway","residential","road")) %>%
  osmdata_sf()

edi_highway
```


## Load Office data
```{r}
edi_office <- opq(edi_bnd) %>%
  add_osm_feature(key = "office") %>%
  osmdata_sf()

edi_office
```





# Data Preprocessing
## Plot buildings as polygons
```{r}
 edi_buildings <- edi_buildings$osm_polygons #choice of polygon was for the area coverage
 edi_buildings <- edi_buildings %>% select(osm_id)
 
 plot(edi_buildings$geometry)
```


## Plot amenities as points
```{r}
edi_amenities <- edi_amenities$osm_points

#select only the columns needed
edi_amenities <- edi_amenities %>%
  select(name, amenity)

#remove NULL values
edi_amenities <- edi_amenities %>%
  filter(!is.na(amenity))

plot(edi_amenities$geometry)

```

## Plot natural as points
```{r}

 edi_natural <- edi_natural$osm_points 
 edi_natural <- edi_natural %>% select(osm_id)
 
 plot(edi_natural$geometry)

```



## Plot Intersections as points: Plot highway as lines and remove null values

```{r}

#First remove all the null values with zero
edi_highway <- edi_highway$osm_lines

edi_highway <- edi_highway %>%
  select(name, access)

edi_highway <- edi_highway %>%
  filter(!is.na(access)) #remove NULL values



#break road network at intersecting lines
rnet_vertices = rnet_breakup_vertices(edi_highway)
 
boundaries = line2points(rnet_vertices) #extract point of intersection

boundaries_df = as.data.frame(st_coordinates(boundaries)) #get points coordinates

boundaries_n = dplyr::summarise(dplyr::group_by(boundaries_df, X, Y), n = dplyr::n()) #group by coordinates to find road segments sharing the same intersection

junction_df = boundaries_n[boundaries_n$n >= 3, ] #Define junctions as points shared among more than 3 street segments

edi_intersections = sf::st_as_sf(junction_df[,1:2], coords = c("X", "Y"), crs = sf::st_crs(edi_highway)) #transform a data frame into an sf object 



#Plot of intersections as points

plot(edi_intersections$geometry)
```


## Plot office as points
```{r}
edi_office <- edi_office$osm_points

#select only the columns needed
edi_office <- edi_office %>%
  select(name, "addr:street")

#remove NULL values
edi_office <- edi_office %>%
  filter(!is.na("addr:street"))

plot(edi_office$geometry)
```






# Composite indicator
## transform the indicator
```{r}
buildings <- st_transform(edi_buildings, 27700)
amenities <- st_transform(edi_amenities, 27700)
natural <- st_transform(edi_natural, 27700)
intersections <- st_transform(edi_intersections, 27700)
offices <- st_transform(edi_office, 27700)
```

## Create grid
```{r}

cell_area <- as_units(1, "km^2") # target grid size

grid_spacing <- sqrt(2*cell_area/sqrt(3)) # size of hexagon calculated from area


grid <- st_make_grid(buildings, cellsize = grid_spacing, what = "polygons", square = FALSE) %>% st_sf() 

grid <- grid %>%
  mutate(
    gid = seq(1:nrow(grid))
  )

plot(grid$geometry)
```

## Building area count
```{r}
grid_buildings <- st_intersection(buildings, grid) # to ensure the shapes of buildings divided when crossed by a hexagon to be able to measure the building area falling within each hexagon cell 

grid_buildings <- grid_buildings %>%
  mutate(
    bld_a = st_area(grid_buildings)
  ) 
grid_buildings <- grid_buildings %>%
  st_drop_geometry() %>% #No need for the spatial referent anymore
  group_by(gid) %>%
  summarise(
    bld_a = sum(bld_a)
  )
``` 



## Intersection counts
```{r}
grid_intersections <- st_intersection(grid, intersections) %>% #duplicate the hexagon cells for every intersecting point
  st_drop_geometry() %>% 
  group_by(gid) %>% #group the hexagons by id to return to a grid with one row per id
  summarise(
    its_c = n() # count the number of rows that had the same gid
  )  

```



## Amenities count
```{r}
grid_amenities <- st_intersection(grid, amenities) %>% # this duplicates the hexagon cells for every intersecting point
  st_drop_geometry() %>% 
  group_by(gid) %>% #this groups the hexagons by id to return to a grid with one row per id
  summarise(
    amt_c = n() #this counts the number of rows that had the same gid
  ) 
```


## Natural counts
```{r}
grid_natural <- st_intersection(grid, natural) %>%
  st_drop_geometry() %>% 
  group_by(gid) %>% 
  summarise(
   ntn_c = n() 
  ) 
```

### Office count
```{r}
grid_office <- st_intersection(grid, offices) %>% 
  st_drop_geometry() %>% 
  group_by(gid) %>% 
  summarise(
    off_c = n() #this counts the number of rows that had the same gid
  ) 
```





# Exploratory Data Analysis
## Histograms
### Histogram of intersections
```{r}
#intersection count
grid %>%
  st_drop_geometry() %>%
  ggplot(aes(x = its_c))+
  geom_histogram(bins = 10)
```

### Histogram of buildings
```{r}
#building area

grid %>%
  ggplot(aes(x = bld_a))+
  geom_histogram(bins = 10)
```

### Histogram of Amenities
```{r}
grid %>%
  ggplot(aes(x = amt_c))+
  geom_histogram(bins = 10)
```


### Histogram of Natural
```{r}
grid %>%
  ggplot(aes(x = ntn_c))+
  geom_histogram(bins = 10)

```


### Histogram of Offices
```{r}
grid %>%
  ggplot(aes(x = off_c))+
  geom_histogram(bins = 10)

```



## Correlation plot
```{r}
#first we need to compute the correlation matrix
corr <- round(cor(st_drop_geometry(grid[,2:6])), 2) 
corr

#then we can compute the p-values - a correlation to be statistically significant needs to have a p-value < 0.001
p.mat <- cor_pmat(st_drop_geometry(grid[,2:6]))
p.mat

#finally we can make the corrplot

ggcorrplot(corr, hc.order = TRUE,
    type = "lower", p.mat = p.mat)
```


## Mapping
```{r}
# I mapped the variables together
tmap_mode("view")
tm_shape(grid) +
    tm_polygons(c("its_c", "bld_a", "amt_c", "ntn_c", "off_c"), alpha = 0.5, style = "equal") +
    tm_facets(sync = TRUE, ncol = 5)
```

#Data Standardisation
## Create a table of all the indicators, replace null values with zero and standardise the data
```{r}

#Create grid of all the indicators with join
grid <- grid %>%
  full_join(grid_intersections, by = c("gid" = "gid")) %>%
  full_join(grid_buildings, by = c("gid" = "gid")) %>%
  full_join(grid_amenities, by = c("gid" = "gid")) %>%
  full_join(grid_natural, by = c("gid" = "gid")) %>%
  full_join(grid_office, by = c("gid" = "gid"))

#Replace null values with zero
grid <- grid %>%
  filter(!is.na(bld_a)) %>%
  replace(is.na(.), 0)
grid <- grid %>%
  filter(!is.na(amt_c)) %>%
  replace(is.na(.), 0)
grid <- grid %>%
  filter(!is.na(its_c)) %>%
  replace(is.na(.), 0)
grid <- grid %>%
  filter(!is.na(ntn_c)) %>%
  replace(is.na(.), 0)
grid <- grid %>%
  filter(!is.na(off_c)) %>%
  replace(is.na(.), 0)

view(grid)


grid <- grid %>%
  mutate( #the scale function returns a matrix while mutate wants a vector, which is why the scale function is within the as.vector function
    its_c = as.vector(scale(its_c)),
    bld_a = as.vector(scale(bld_a)),
    amt_c = as.vector(scale(amt_c)),
    ntn_c = as.vector(scale(ntn_c)),
    off_c = as.vector(scale(off_c))
  )


summary(grid[,2:6]) #tranformed the mean to 0

sapply(grid[2:6], sd) #transformed the standard deviation to 1

```



## Combine the indicators
```{r}
# I am giving priority to the amenities, nature and office
walkability <- grid %>%
  mutate(
    walkability_0 = 4*amt_c + bld_a + its_c + 3*ntn_c + 2*off_c
  ) #I prioritised amenities, natural and offices, by giving them more weights than building area and intersections.
  
```


## Sensitivity Analysis
```{r}
#create different composite indicators with different weights
walkability <- walkability %>%
  mutate(
    walkability_1 = 0.45*amt_c + bld_a + 0.32*its_c + 0.2*ntn_c + off_c,
    walkability_2 = amt_c + bld_a + its_c + ntn_c + off_c
  )

#compute correlation between different composite indicators 

walkability %>%
  st_drop_geometry() %>%
  select(starts_with("walkability")) %>%
  cor()
```

## Visualisation
```{r}

tmap_mode("view")
tm_shape(walkability) +
    tm_fill("walkability_2", alpha = 0.5, palette = "viridis", midpoint = 0) #since there was not so much difference, I used the walkability_2, which had same weight for all the variables
```




#Service Analysis
## Data collection
```{r}

#I listed services that could be accesses within 5, 10 and 15 minutes
edi_services <- opq(edi_bnd) %>%
  add_osm_feature(key = "amenity", value = c("cafe","fast_food", "exhibition_centre", "post_office", "community_centre","casino","cinema","atm", "kindergarten", "clinic")) %>%
  osmdata_sf()

edi_services <- edi_services$osm_points
```


## transform the grid (walkability to WGS84)
```{r}
grid <- walkability %>% st_transform(4326)
amt <- edi_amenities %>% st_transform(4326)

grid_centroids <- st_centroid(grid)
plot(grid_centroids$geom)

```




# Site Selection
```{r}

tmap_mode("view")
  tm_shape(grid_centroids) + tm_dots("walkability_2", alpha = 0.5, palette = "RdYlGn", midpoint = 0)
  
```

## Check for the services available at the site with the highest walkability
```{r}
#I chose two sites that represents high and low walkability score

site1 <- grid_centroids %>%
  filter(gid == "353") #Site 353 is among the area with highest walkability scores


site2 <- grid_centroids %>%
  filter(gid == "422") # Site 422 is among the area on the map with lowest walkability scores

```



## isochrones
```{r}

iso <- osrmIsochrone(loc = site1, 
                     breaks = c(5,10,15),
                     osrm.profile = "foot") #Isochrone for the the high walkability site
iso <- iso %>% st_make_valid() 


tmap_mode("view")
  tm_shape(site1) +
    tm_dots(col = "red")+
    tm_shape(iso) +
    tm_polygons("isomax", alpha = 0.5, style = "cat") 

```



```{r}
iso2 <- osrmIsochrone(loc = site2, 
                     breaks = c(5,10,15),
                     osrm.profile = "foot") #Isochrone for the low walkability site
iso2 <- iso2 %>% st_make_valid() 


tmap_mode("view")
  tm_shape(site2) +
    tm_dots(col = "red")+
    tm_shape(iso2) +
    tm_polygons("isomax", alpha = 0.5, style = "cat") 
```


### Service accessibility for for Site1 within 5, 10 and 15 minutes
```{r}
#intersect the edi_services with the isochrones
#library(sf)
services <- edi_services

site_services <- st_intersection(iso, services) %>%
  st_drop_geometry() %>% # spatial data is not necessary, hence I chose to summarise by grouping
  group_by(id, amenity) %>%
  summarise(
    srv_c = n() # Count the number of the amenities that intersect with isochrones
  )

st_crs(iso)

view(site_services)

```

### Service accessibility for for Site2 within 5, 10 and 15 minutes
```{r}
#intersect the edi_services with the isochrones
#library(sf)
services <- edi_services

site_services2 <- st_intersection(iso2, services) %>%
  st_drop_geometry() %>% # spatial data is not necessary, hence I chose to summarise by grouping
  group_by(id, amenity) %>%
  summarise(
    srv_c = n() # Count the number of the amenities that intersect with isochrones
  )

st_crs(iso2)

view(site_services2)

```

### Check the max and min in the dataset
```{r}
summary(site_services)
```

```{r}
summary(site_services2)
```

### Map the result
```{r}
# ste the site_services as the sf class to be able to map
site_services <- site_services %>%
  pivot_wider(id_cols = "id", names_from = "amenity", values_from = "srv_c") %>%
  inner_join(iso, by = c("id" = "id")) %>%
  st_as_sf()

tmap_mode("view")
  tm_shape(site_services) + 
    tm_fill(col = c("cafe","fast_food"), palette = "viridis", breaks = c(1, 20, 40, 60, 80, 100)) #set the breaks to be able to compare the maps between each other taking into consideration the count max and min 
```


```{r}
# the site_services as the sf class to be able to map
site_services2 <- site_services2 %>%
  pivot_wider(id_cols = "id", names_from = "amenity", values_from = "srv_c") %>%
  inner_join(iso2, by = c("id" = "id")) %>%
  st_as_sf()

tmap_mode("view")
  tm_shape(site_services2) + 
    tm_fill(col = c("cafe","fast_food"), palette = "viridis", breaks = c(1, 20, 40, 60, 80, 100)) #set the breaks to be able to compare the maps between each other taking into consideration the count max and min 
```
